---
title: "PA1_template"
author: "Yair Meidan"
date: "Wednesday, April 15, 2015"
output: html_document
---

## Initializing the environment

1. Set global options
- Make sure that the code for generating the output is always present
- Make sure that results are always shown
- Make sure that warnings and messages (e.g. while loading packages) are not displayed
```{r setoptions, echo="TRUE"}
knitr::opts_chunk$set(
        echo=TRUE
        ,results="show"
        ,warning=FALSE
        ,message=FALSE
        ,hard_wrap=FALSE
)
```

2. Remove all objects from memory  (clear it)
```{r clear_memory}
rm(list=ls())
```

3. Install required packages (only if needed) and load them
```{r}
ipc<-as.data.frame(installed.packages())
if(!("dplyr" %in% ipc$Package)){install.packages("dplyr")}
if(!("chron" %in% ipc$Package)){install.packages("chron")}
if(!("ggplot2" %in% ipc$Package)){install.packages("ggplot2")}
if(!("xtable" %in% ipc$Package)){install.packages("xtable")}
if(!("lubridate" %in% ipc$Package)){install.packages("lubridate")}
library(dplyr)
library(chron) 
library(ggplot2)
library(xtable)
library(lubridate)
```

## Loading and preprocessing the data

1. Load the data
- download zip file only if needed
- unzip and read entire csv file
- assume that the user has already defined a working directory with setwd()
```{r}
# download zip file if needed 
if (!file.exists("repdata-Fdata-Factivity.zip")) {
        fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileUrl,destfile="repdata-Fdata-Factivity.zip")
}

# read data from zip file
d_original<-read.csv(unz("repdata-Fdata-Factivity.zip", "activity.csv"))
```

2. Process/transform the data (if necessary) into a format suitable for analysis
- make a copy of the original data, for further processing
```{r}
# make a copy of original data
d<-d_original

# format as date
d$date<-as.Date(d$date)

# make a data set where all steps are not null
d_not_null<-d[!is.na(d$steps),]
```

## What is mean total number of steps taken per day?

For this part of the assignment, missing values in the dataset are ignored.

1. Calculate the total number of steps taken per day
```{r}
# total daily steps
daily_steps<-aggregate(
        x=d_not_null$steps
        ,by=list(date=d_not_null$date)
        ,FUN=sum
)

# rename
names(daily_steps)[names(daily_steps)=="x"] <- "total_steps"
```

2. Make a histogram of the total number of steps taken each day
```{r}
ggplot(
        data = daily_steps
        ,aes(x=total_steps)
)+ 
        geom_histogram(
                binwidth=2500
                ,colour="black"
                , fill="grey"
        )+
        xlab("Total daily steps")+
        ylab("Number of days")+
        theme(text = element_text(size=20))+
        ggtitle(expression(atop("Total number of steps taken per day", atop(italic("missing values have been removed"), ""))))
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
# calculate, convert scientific to decimal format
daily_steps_mean<-format(round(mean(daily_steps$total_steps), 2),2)
daily_steps_median<-format(round(median(daily_steps$total_steps), 2),2)
```
- The mean of the total number of steps taken per day is `r daily_steps_mean`.
- The median of the total number of steps taken per day is `r daily_steps_median`.

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
# transform interval format into HH:MM
d_not_null$interval_hhmm<-ifelse(
        d_not_null$interval<10
        ,paste("00:0",as.character(d_not_null$interval),":00",sep="")
        ,ifelse(
                d_not_null$interval<100
                ,paste("00:",as.character(d_not_null$interval),":00",sep="")
                ,ifelse(
                        d_not_null$interval<1000
                        ,paste(
                                "0"
                                ,substr(as.character(d_not_null$interval),1,1)
                                ,":"
                                ,substr(as.character(d_not_null$interval),2,3)
                                ,":00"
                                ,sep="")
                        ,paste(
                                substr(as.character(d_not_null$interval),1,2)
                                ,":"
                                ,substr(as.character(d_not_null$interval),3,4)
                                ,":00"
                                ,sep="")   
                )
        )
)

# time format
d_not_null$interval_time<-strftime(as.POSIXlt(d_not_null$interval_hhmm,format="%H:%M"))

# mean steps per interval
interval_steps<-aggregate(
        x=d_not_null$steps
        ,by=list(interval=d_not_null$interval_hhmm)
        ,FUN=mean
)

# rename
names(interval_steps)[names(interval_steps)=="x"] <- "average_steps"

# produce plot
qplot(
        interval
        ,average_steps
        ,data=interval_steps
        ,geom=c("line","point")
        ,group = 1
        ,xlab="Interval"
        ,ylab="Total steps"
)+ 
        scale_x_discrete(breaks=c("0:00:00","06:00:00","12:00:00","18:00:00"))+
        theme(text = element_text(size=20))+
        ggtitle(expression(atop("Number of steps taken per interval", atop(italic("averaged across all days, missing values have been removed"), ""))))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# maximum value
mx<-max(interval_steps$average_steps)

# interval of maximum value
mx_interval<-interval_steps[interval_steps$average_steps==mx,"interval"]

# scientific to decimal formatting
mx<-format(round(mx,2),2)

```
- The maximum average number of steps is `r mx`.
- It is contained in the `r mx_interval` interval.


## Imputing missing values

- Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
# missing values per variable
NA_steps<-sum(is.na(d$steps))
NA_date<-sum(is.na(d$date))
NA_interval<-sum(is.na(d$interval))

# sum of missing values
Na_total<-NA_steps+NA_date+NA_interval
```
- the total number of missing values in the dataset is `r Na_total`.


2. Fill in all of the missing values in the dataset by replacing them with the median for their 5-minute interval.
```{r}
# calculate interval medians
interval_median_steps<-aggregate(
        x=d_not_null$steps
        ,by=list(interval=d_not_null$interval)
        ,FUN=median
)

# join original data with medians
original_d_with_medians<-merge(
        x=d
        ,y=interval_median_steps
        ,by.x="interval"
        ,by.y="interval"
        ,all.x=TRUE
)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# same structure, replaced values
d_with_replaced_NAs<-data.frame(
        interval=original_d_with_medians$interval
        ,steps=ifelse(
                is.na(original_d_with_medians$steps)
                ,original_d_with_medians$x
                ,original_d_with_medians$steps
        )
        ,date=original_d_with_medians$date
)
```

4. Make a histogram of the total number of steps taken each day. 
```{r}
# compute daily steps
daily_steps_with_replaced_NAs<-aggregate(
        x=d_with_replaced_NAs$steps
        ,by=list(date=d_with_replaced_NAs$date)
        ,FUN=sum
)

# rename
names(daily_steps_with_replaced_NAs)[names(daily_steps_with_replaced_NAs)=="x"] <- "total_steps_after_replacing_NAs"

# plot the histogram
ggplot(
        data = daily_steps_with_replaced_NAs
        ,aes(x=total_steps_after_replacing_NAs)
)+ 
        geom_histogram(
                binwidth=2500
                ,colour="black"
                , fill="grey"
        )+
        xlab("Total daily steps")+
        ylab("Number of days")+
        theme(text = element_text(size=20))+
        ggtitle(expression(atop("Total number of steps taken per day", atop(italic("missing values have been replaced with interval median"), ""))))
```

- Calculate and report the mean and median total number of steps taken per day.
- Do these values differ from the estimates from the first part of the assignment? 
```{r}
# measures for replaced missing values
daily_steps_mean_with_replaced_NAs<-round(mean(daily_steps_with_replaced_NAs$total_steps),2)
daily_steps_median_with_replaced_NAs<-median(daily_steps_with_replaced_NAs$total_steps)

daily_steps_mean_with_replaced_NAs_to_print<-format(daily_steps_mean_with_replaced_NAs,2)

# measures for omitted missing values
daily_steps_mean<-round(mean(daily_steps$total_steps),2)
daily_steps_median<-median(daily_steps$total_steps)

# difference in measures
means_diff<-format(round((daily_steps_mean_with_replaced_NAs-daily_steps_mean),2),2)
medians_diff<-(daily_steps_median_with_replaced_NAs-daily_steps_median)
```
- When missing values are replaced with interval medians, the mean of the total number of steps taken per day is `r daily_steps_mean_with_replaced_NAs_to_print`, which differs by `r means_diff` from the case when missing values are removed.
- When missing values are replaced with interval medians, the median of the total number of steps taken per day is `r daily_steps_median_with_replaced_NAs`, which differs by `r medians_diff` from the case when missing values are removed.

- What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# data frame to examine the effect of replacing missing values
daily_steps_with_and_without_replacing_NAs<-rbind(
        data.frame(
                total_steps=daily_steps$total_steps
                ,NAs_replaced="no"
        )        
        ,data.frame(
                total_steps=daily_steps_with_replaced_NAs$total_steps_after_replacing_NAs
                ,NAs_replaced="yes"
        )        
)

# comparative plot - 2 histograms
ggplot(
        data = daily_steps_with_and_without_replacing_NAs
        ,aes(x=total_steps,fill=NAs_replaced)
)+ 
        geom_bar(
                position="dodge"
                ,binwidth=500
        )+
        xlab("Total daily steps")+
        ylab("Number of days")+
        theme(text = element_text(size=20))+
        ggtitle(expression(atop("Total number of steps taken per day", atop(italic("with and without replacing missing values with medians"), ""))))
``` 

- In terms of central measures, the mean and median daily steps have decreased after missing values had been replaced.
- In terms of distribution shape, the above comparative histograms demonstrate how imputing missing data on the estimates of the total daily number of steps has almost no effect, except for a very specific region.
- That specific region of difference needs further examination

```{r xtable, results="asis"}
# original data for days with missing values only
d_null<-d[is.na(d$steps),]

# number of days with missing values
dates_with_NA<-aggregate(
        x=d_null$steps
        ,by=list(date=d_null$date)
        ,FUN=length
) # --> each interval misses exactly 288 values of steps

# rename
names(dates_with_NA)[names(dates_with_NA)=="x"] <- "number_of_missing_values"

# make sure the date is printed correctly
xt<-xtable(data.frame(
             date=as.character(dates_with_NA$date)
             ,number_of_missing_values=dates_with_NA$number_of_missing_values
        )
)
print(xt, type="html")
```
- It seems that only 8 days are the source of difference, each with exactly 288 missing values for interval steps.
- Here is a code for closer look on these 8 days.

```{r xtable2, results="asis"}
# daily steps in dates where missing values were replaced
total_daily_steps_where_replaced<-aggregate(
        x=d_with_replaced_NAs[d_with_replaced_NAs$date %in% dates_with_NA$date,"steps"]
        ,by=list(d_with_replaced_NAs[d_with_replaced_NAs$date %in% dates_with_NA$date,"date"])
        ,FUN=sum
)

# rename variables
names(total_daily_steps_where_replaced)[names(total_daily_steps_where_replaced)=="Group.1"] <- "date"
names(total_daily_steps_where_replaced)[names(total_daily_steps_where_replaced)=="x"] <- "total_steps_after_replacing_NAs"

total_daily_steps_where_replaced$date<-as.Date(total_daily_steps_where_replaced$date, format="%Y-%m-%d")

# calculate region of difference
N_steps_in_replaced_days<-unique(total_daily_steps_where_replaced$total_steps_after_replacing_NAs)

# xt2<-xtable(total_daily_steps_where_replaced)
# make sure the date is printed correctly
xt2<-xtable(data.frame(
             date=as.character(total_daily_steps_where_replaced$date)
             ,total_steps_after_replacing_NAs=total_daily_steps_where_replaced$total_steps_after_replacing_NAs
        )
)
print(xt2, type="html")
```
- Further examination revealed that exactly `r N_steps_in_replaced_days` daily steps were calculated for each of these days.
- This is the only column in the comparative histogram that changed when missing values were imputed with interval medians.

## Are there differences in activity patterns between weekdays and weekends?

- Note: For this part the dataset used in the one with the filled-in missing values.

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
# save original locale, to reset later
original_locale<-Sys.getlocale("LC_TIME")

# set locale to English
Sys.setlocale("LC_TIME", "English")

# create factor variable
d_with_replaced_NAs$weekpart<-ifelse(
        weekdays(d_with_replaced_NAs$date) %in% c("Saturday","Sunday")    
        ,"weekend"
        ,"weekday"
)
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r}
# Calculate the average number of steps taken per interval
interval_weekpart_steps<-aggregate(
        x=d_with_replaced_NAs$steps
        ,by=list(interval=d_with_replaced_NAs$interval, weekpart=d_with_replaced_NAs$weekpart)
        ,FUN=mean
)

# rename variable
names(interval_weekpart_steps)[names(interval_weekpart_steps)=="x"] <- "average_steps"

# produce plot
qplot(
        interval
        ,average_steps
        ,data=interval_weekpart_steps
        ,geom=c("line","point")
        ,group = 1
        ,xlab="Interval"
        ,ylab="Number of steps"
        ,facets = weekpart~.
)+ 
        theme(text = element_text(size=20))+
        ggtitle(expression(atop("Number of steps taken per interval", atop(italic("avg. across all days, missing values replaced with medians"), ""))))

# reset locale to original
Sys.setlocale("LC_TIME", original_locale)
```